---
title: "stat3355_finalProject"
author: "Muhammad Zubair"
date: "5/7/2022"
output: pdf_document
---

# Libraries
```{r}
library(tidyr)
library(readr)
library(dplyr)
library(stringr)
library(ggplot2)
library(ggcorrplot)
library(ggdark)
library(scales)
library(car)
library(MASS)
library(DAAG)
library(lubridate)
```


# Data Cleaning
```{r}
# Read in the data frame
df <- read_csv("Metro_Interstate_Traffic_Volume.csv.gz")
```

## Data Cleaning - Date & Time
```{r}
# Separating the date_time into date and time column
df$date <- sapply(strsplit(as.character(df$date_time)," "), "[", 1)
df$date <- as.Date(df$date, format="%Y-%m-%d")
df$time <- sapply(strsplit(as.character(df$date_time)," "), "[", 2)

df$Date <- as.Date(df$date_time)
df$Time <- format(as.POSIXct(df$date_time), format = "%H:%M:%S")

df$TimeSec <- as.numeric(as.POSIXct(df$Time, format = "%H:%M:%S")) %% 86400

summary(df$temp)
df$temp[which(df$temp < 5)] <- NA
summary(df$temp)

hist(df$traffic_volume, main = "Histogram of Traffic Volume", xlab = "Traffic Volume")
```

## Data Cleaning - Holiday
```{r}
# Mark holiday as a logical value
df$holiday <- ifelse(df$holiday == "None", "None", "Yes")

# Table of holidays before cleaning

table(df$holiday)

# Now I can look at the data where holiday = yes, and make the whole day as yes
# Pull the dates of each holiday
df_holiday <- subset(df, holiday=='Yes')
df_holiday <- subset(df, select = c(holiday, date))

# Marking entire day as holiday, instead of only first hour

# Holidays in 2013
df$holiday[format(df$date, '%Y-%m-%d') %in% '2012-10-08'] <- "Yes"
df$holiday[format(df$date, '%Y-%m-%d') %in% '2012-11-12'] <- "Yes"
df$holiday[format(df$date, '%Y-%m-%d') %in% '2012-11-22'] <- "Yes"
df$holiday[format(df$date, '%Y-%m-%d') %in% '2012-12-25'] <- "Yes"
df$holiday[format(df$date, '%Y-%m-%d') %in% '2013-01-01'] <- "Yes"
df$holiday[format(df$date, '%Y-%m-%d') %in% '2013-02-18'] <- "Yes"
df$holiday[format(df$date, '%Y-%m-%d') %in% '2013-05-27'] <- "Yes"
df$holiday[format(df$date, '%Y-%m-%d') %in% '2013-07-04'] <- "Yes"
df$holiday[format(df$date, '%Y-%m-%d') %in% '2013-09-02'] <- "Yes"
df$holiday[format(df$date, '%Y-%m-%d') %in% '2013-10-14'] <- "Yes"
df$holiday[format(df$date, '%Y-%m-%d') %in% '2013-11-11'] <- "Yes"
df$holiday[format(df$date, '%Y-%m-%d') %in% '2013-11-28'] <- "Yes"
df$holiday[format(df$date, '%Y-%m-%d') %in% '2013-12-25'] <- "Yes"

# Holidays in 2014
df$holiday[format(df$date, '%Y-%m-%d') %in% '2014-01-01'] <- "Yes"
df$holiday[format(df$date, '%Y-%m-%d') %in% '2014-01-20'] <- "Yes"
df$holiday[format(df$date, '%Y-%m-%d') %in% '2014-02-17'] <- "Yes"
df$holiday[format(df$date, '%Y-%m-%d') %in% '2014-05-26'] <- "Yes"

# Holidays in 2015
df$holiday[format(df$date, '%Y-%m-%d') %in% '2015-07-03'] <- "Yes"
df$holiday[format(df$date, '%Y-%m-%d') %in% '2015-08-27'] <- "Yes"
df$holiday[format(df$date, '%Y-%m-%d') %in% '2015-09-07'] <- "Yes"
df$holiday[format(df$date, '%Y-%m-%d') %in% '2015-10-12'] <- "Yes"
df$holiday[format(df$date, '%Y-%m-%d') %in% '2015-11-11'] <- "Yes"
df$holiday[format(df$date, '%Y-%m-%d') %in% '2015-11-26'] <- "Yes"
df$holiday[format(df$date, '%Y-%m-%d') %in% '2015-12-25'] <- "Yes"

# Holidays in 2016
df$holiday[format(df$date, '%Y-%m-%d') %in% '2016-01-01'] <- "Yes"
df$holiday[format(df$date, '%Y-%m-%d') %in% '2016-02-15'] <- "Yes"
df$holiday[format(df$date, '%Y-%m-%d') %in% '2016-05-30'] <- "Yes"
df$holiday[format(df$date, '%Y-%m-%d') %in% '2016-07-04'] <- "Yes"
df$holiday[format(df$date, '%Y-%m-%d') %in% '2016-08-25'] <- "Yes"
df$holiday[format(df$date, '%Y-%m-%d') %in% '2016-09-05'] <- "Yes"
df$holiday[format(df$date, '%Y-%m-%d') %in% '2016-10-10'] <- "Yes"
df$holiday[format(df$date, '%Y-%m-%d') %in% '2016-11-11'] <- "Yes"
df$holiday[format(df$date, '%Y-%m-%d') %in% '2016-11-24'] <- "Yes"
df$holiday[format(df$date, '%Y-%m-%d') %in% '2016-12-26'] <- "Yes"

# Holidays in 2017
df$holiday[format(df$date, '%Y-%m-%d') %in% '2017-01-02'] <- "Yes"
df$holiday[format(df$date, '%Y-%m-%d') %in% '2017-01-16'] <- "Yes"
df$holiday[format(df$date, '%Y-%m-%d') %in% '2017-02-20'] <- "Yes"
df$holiday[format(df$date, '%Y-%m-%d') %in% '2017-05-29'] <- "Yes"
df$holiday[format(df$date, '%Y-%m-%d') %in% '2017-07-04'] <- "Yes"
df$holiday[format(df$date, '%Y-%m-%d') %in% '2017-08-24'] <- "Yes"
df$holiday[format(df$date, '%Y-%m-%d') %in% '2017-09-04'] <- "Yes"
df$holiday[format(df$date, '%Y-%m-%d') %in% '2017-10-09'] <- "Yes"
df$holiday[format(df$date, '%Y-%m-%d') %in% '2017-11-10'] <- "Yes"
df$holiday[format(df$date, '%Y-%m-%d') %in% '2017-11-23'] <- "Yes"
df$holiday[format(df$date, '%Y-%m-%d') %in% '2017-12-25'] <- "Yes"

# Holidays in 2018
df$holiday[format(df$date, '%Y-%m-%d') %in% '2018-01-01'] <- "Yes"
df$holiday[format(df$date, '%Y-%m-%d') %in% '2018-01-15'] <- "Yes"
df$holiday[format(df$date, '%Y-%m-%d') %in% '2018-02-19'] <- "Yes"
df$holiday[format(df$date, '%Y-%m-%d') %in% '2018-05-28'] <- "Yes"
df$holiday[format(df$date, '%Y-%m-%d') %in% '2018-07-04'] <- "Yes"
df$holiday[format(df$date, '%Y-%m-%d') %in% '2018-08-23'] <- "Yes"
df$holiday[format(df$date, '%Y-%m-%d') %in% '2018-09-03'] <- "Yes"

# Checking to see if values were added
table(df$holiday)

# Making a copy of df for future data cleaning purposes
df_holiday_cleaned <- data.frame(df)
```

## Data Cleaning - Weather frequency
```{r}
# Identify which types of weather occur the most
df_max_occur <- df_holiday_cleaned %>% 
    gather("key", "value", weather_description) %>%
    group_by(date) %>% 
    dplyr::summarise(max.occurence = names(which.max(table(value))))

head(df_max_occur)

# Merging the df of max occurrences with the original df
df_max_occur$date <- as.Date(df_max_occur$date, format="%Y-%m-%d")
df_clean <- full_join(df_holiday_cleaned, df_max_occur, 
                       by = c("date" = "date"))
```

## Data Cleaning - Traffic volume intervals
```{r}
# Create intervals of traffic volume for easier comparison
df_clean$traffic_interval_of_250 <- floor(df_clean$traffic_volume / 250)+1
df_clean$traffic_interval_of_250 <- as.factor(df_clean$traffic_interval_of_250)
```

## Data Cleaning - Cleaned data frame
```{r}
df_clean$max.occurence <- as.factor(df_clean$max.occurence)
df_clean$holiday <- as.factor(df_clean$holiday)
df_clean$temp <- df_clean$temp - 273.15

summary(df_clean)

```



# Data Visualization
```{r}
# Create a data frame for data visualizations
df_visualization <- subset(df_clean, 
                            select = -c(max.occurence, weather_description, date_time))
df_visualization$holiday <- ifelse(df_visualization$holiday == "Yes", 1, 0)
df_visualization$weather_main <- as.factor(df_visualization$weather_main)

str(df_visualization)
```



## Visualization - Time series
```{R}
df_visualization$time <- as.POSIXct(df_visualization$time, format="%H:%M:%S")
df_visualization$traffic_interval_of_250 <- as.integer(df_visualization$traffic_interval_of_250)
```

## doesn't work independently 
## Visualization - Correlation matrix
```{R}
df_cor_mat <- subset(df_visualization, !is.na(temp),
              select = -c(date, traffic_volume))

df_cor_mat$rain_1h <- ifelse(df_cor_mat$rain_1h == 0, 0, 1)
df_cor_mat$snow_1h <- ifelse(df_cor_mat$snow_1h == 0, 0, 1)
df_cor_mat$time <- sapply(strsplit(as.character(df_cor_mat$time)," "), "[", 2)
df_cor_mat$time <- sapply(strsplit(as.character(df_cor_mat$time),":"), "[", 1)
df_cor_mat$time <- as.numeric(df_cor_mat$time)

df_cor_mat <- df_cor_mat[sapply(df_cor_mat,is.numeric)]
corr <- cor(df_cor_mat)
ggcorrplot(corr, lab = TRUE, hc.order = TRUE, type = "lower", 
          outline.color = "white",
          colors = c("white", "lightblue", "blue")) + 
theme(axis.text.x = element_text(angle = 90, hjust=1))
```

## Visualization - Histogram of Response
```{r}
ggplot(df_visualization, aes(x=time, y=traffic_interval_of_250))+
  geom_bar(size=1.5, stat = "identity", fill="LIGHTBLUE") +
  labs(x=NULL, y="Traffic volume", title = "Traffic Volume by Hour") + 
  scale_x_datetime(breaks = date_breaks("2 hour"), 
                    labels=date_format("%H:%M:%S"))+
  theme(axis.text.x = element_text(angle = 45, hjust=1)) + 
  theme(plot.title = element_text(hjust = 0.5)) 
```

## Visualization - Temperature
```{r}
ggplot(subset(df_clean, temp > -50), aes(x = temp, y = traffic_volume))+
  geom_point(color = "lightblue", cex = 0.5)+
  geom_smooth(method = lm, formula = y~ x, se = F, color = "blue")+
  labs(x = "Temperature in Celsius", y = "Traffic Volume", title = "Temperature and Traffic Volume")

# 
# df_clean$timeGroup <- df_clean$time
# df_clean$timeGroup <- sapply(strsplit(as.character(df_clean$time)," "), "[", 2)
# df_clean$timeGroup <- sapply(strsplit(as.character(df_clean$time),":"), "[", 1)
# as.numeric(df_clean$timeGroup)
# df_clean$timeGroup <- with(df_clean,  ifelse(time >= 5 & time<=8, "earlymorning",
#                             ifelse(time>8 & time<=11, "morning",
#                             ifelse(time > 11 & time<=15, "afternoon" , 
#                             ifelse(time > 15 & time<=20, "evening", "night")))))
# table(df_clean$timeGroup)

ggplot(subset(df_clean, temp > -50), aes(y = temp, x = time, color = traffic_volume))+
  geom_jitter()+
  labs(x = "Time", y = "Temperature", title = "Time, Temperature, and Traffic Volume")+
  theme(axis.text.x = element_text(angle = 90))
  

```

## Visualization - Hourly Rain
```{r}
ggplot(subset(df_clean, rain_1h < 10), aes(x = rain_1h, y = traffic_volume))+
  geom_point(color = "lightblue", cex = 1)+
  geom_smooth(method = lm, formula = y ~ x, se = F, color = "blue")+
  labs(x = "Rain per hour in mm", y = "Traffic Volume", title = "Hourly Rain and Traffic Volume")

```

## Visualization - Hourly Snow
```{r}
ggplot(df_clean, aes(x = snow_1h, y = traffic_volume))+
  geom_point(color = "lightblue", cex = 1)+
  geom_smooth(method = lm, formula = y ~ x, se = F, color = "blue")+
  labs(x = "Snow per hour in mm", y = "Traffic Volume", title = "Hourly Snow and Traffic Volume")
```

## Visualization - Cloudcover
```{r}
ggplot(df_clean, aes(x = clouds_all, y = traffic_volume))+
  geom_jitter(cex = 0.5, color = "lightblue")+
  geom_smooth(method = lm, formula = y ~ x, se = F, color = "blue")+
  labs(x = "Total cloudcover percentage", y = "Traffic Volume", title = "Cloudcover and Traffic Volume")

```

## Visualization - Holiday
```{r}
ggplot(df_clean, aes(x = holiday, y = traffic_volume, fill = holiday))+
  geom_boxplot(color = "black")+
  labs(x = "Holiday or Not", y = "Traffic Volume", title = "Traffic Volume on Holiday vs Regular Days")+
  geom_hline(yintercept = 3380, linetype = "dashed", color = "black")
```

## Visualization - Weather Main
```{r}
ggplot(df_clean, aes(x = weather_main, y = traffic_volume, fill = weather_main))+
  geom_boxplot(color = "black")+
  labs(x = "Primary Weather Description", y = "Traffic Volume", 
       title = "Overall Weather and Traffic Volume")+
  geom_hline(yintercept = 3380, linetype = "dashed", color = "black")
```

# Primary Model

## Primary - Time intervals
```{r}
df_time_factor <- df_visualization

# df_time_factor$groups <- cut(df_time_factor$TimeSec,
#                              breaks = c(-1, 28800, 39600, 54000, 68400, Inf),
#                              labels = c("Night", "Early Morning", "Morning", "Afternoon", "Evening"))
df_time_factor$groups <- cut(df_time_factor$TimeSec,
                             breaks = c(-1, 21600, 43200, 64800, Inf),
                             labels = c("Night", "Early", "Midday", "Afternoon"))
# df_time_factor$groups <- cut(df_time_factor$TimeSec, 
#                              breaks = c(-1, 28800, 54000, 68400, Inf),        
#                              labels = c("Night", "Morning", "Afternoon", "Evening"))

table(df_time_factor$groups)

df_model <- subset(df_clean,
                   select = -c(date_time, weather_description, date, Date,TimeSec, max.occurence, traffic_interval_of_250))
df_model$timeInterval <- factor(df_time_factor$groups)

summary(df_model)

```

## Primary - Linear model with all regressors
```{r}
fit1 <- lm(data = df_model, traffic_volume~. -time - Time)
summary(fit1)

```

# Reduced model
```{r}
df_reduced1 <- subset(df_model,
                      select = -c( rain_1h, snow_1h, clouds_all))
fit2 <- lm(data = df_reduced1,
           traffic_volume ~. -time - Time)
fit2b <- lm(data = df_reduced1,
            traffic_volume ~. -timeInterval)
summary(fit2)
```

## Reduced - Time Intervals Box
```{r}
ggplot(data = df_reduced1, aes(x = timeInterval, y = traffic_volume))+
  geom_jitter(cex = 0.2, color = "lightgray")+
  geom_boxplot(fill = NA)+
  geom_hline(yintercept = 3380, color = "deepskyblue3")+
  labs(x = "Time of day", y = "Traffic Volume", title = "Boxplot of Traffic Volume by Time")
  
median(df_reduced1$traffic_volume[which(df_reduced1$timeInterval == "Early"
                                        & !is.na(df_reduced1$timeInterval))])
median(df_reduced1$traffic_volume[which(df_reduced1$timeInterval == "Midday")])
median(df_reduced1$traffic_volume[which(df_reduced1$timeInterval == "Afternoon")])
median(df_reduced1$traffic_volume[which(df_reduced1$timeInterval == "Night")])
```

## Reduced - ANOVA table comparison
```{r}
anova(fit2, fit1)
```

## Reduced - VIF table
```{r}
vif(fit2)

```

# Residual and Influence Analysis

## Residuals - Standardized
```{r}
#Standardized residuals
length(which(stdres(fit2) > 3))
length(which(stdres(fit2) < -3))

barplot(height = stdres(fit2),
        main = "Standardized Residuals", xlab = "index", 
        ylab = "Standardized Residual", ylim = c(-4, 4))
abline(h = 3, col = "lightblue")
abline(h = -3, col = "lightblue")
```

## Residuals - Studentized
```{r}
#Studentized residuals
length(which(studres(fit2) > 3))
length(which(studres(fit2) < -3))

barplot(height = studres(fit2),
        main = "Studentized Residuals", xlab = "index",
        ylab = "Studentized Residual", ylim = c(-4, 4))
abline(h = 3, col = "lightblue")
abline(h = -3, col = "lightblue")
```

## Residuals - R Student
```{r}
#R student residuals
cor.qt <- qt(0.05/(2 * 48204), 48176, lower.tail = F)
cor.qt
length(which(rstudent(fit2) > cor.qt))
length(which(rstudent(fit2) < -cor.qt))

barplot(height = rstudent(fit2),
        main = "R Student Residuals", xlab = "index",
        ylab = "R Student Residual", ylim = c(-6, 6))
abline(h=cor.qt, col = "lightblue", lwd=2)
abline(h=-cor.qt, col = "lightblue", lwd=2)
```

## Residuals - Residuals vs fitted, histogram, QQ
```{r}
residualPlot(fit2b, type = "rstudent", quadratic=F, col = "lightblue",
             main = "Residuals vs Fitted Values Pre-Transformation")

hist(studres(fit2), breaks = 25, freq=F, 
     col = "lightblue", 
     main = "Histogram of Residuals Pre-Transformation",
     xlab = "Studentized Residuals")

qqPlot(fit2, col = "grey", col.lines = "cornflowerblue",
       ylab = "Studentized Residuals",
       main = "Quantile Comparison Pre-Transformation")
```

## Residuals - Summary
```{r}
df_ResPre <- df_reduced1[which(stdres(fit2) > 3 |
                                     stdres(fit2) < -3 |
                                     studres(fit2) > 3 |
                                     studres(fit2) < -3 |
                                     rstudent(fit2) > cor.qt |
                                     rstudent(fit2) < -cor.qt), ]
```

## Residuals - box cox
```{r}
df_positive <- df_reduced1
df_positive$traffic_volume <- (df_positive$traffic_volume + 0.01)
fit3 <- lm(data = df_positive, traffic_volume~. )
fit3b <- lm(data = df_positive, traffic_volume~. -time - Time)

boxcox.traffic <- boxCox(fit3)
boxcox.trafficb <- boxCox(fit3b)
boxcox.traffic$x[which.max(boxcox.traffic$y)]

```

# Transformation
```{r}
# Square root transformation on the response
df_transformed <- data.frame(traffic_volume = (df_reduced1[,4])^(0.5),
                             holiday = df_reduced1[,1],
                             temp = df_reduced1[,2],
                             weather_main = df_reduced1[,3],
                             time = df_reduced1[,5], 
                             Time = df_reduced1[,6],
                             timeInterval = df_reduced1[,7])
summary(df_transformed)

fit4 <- lm(data = df_transformed, traffic_volume~. -time -Time)
fit4b <- lm(data = df_transformed, traffic_volume~. -timeInterval)

summary(fit4)
```

# New Residual Analysis Post-Transformation

## New Residuals - Residual plot comparison
```{r}
residualPlot(fit2b, type = "rstudent", quadratic=F, col = "lightblue",
             main = "Before",
             ylim = c(-5, 5))

residualPlot(fit4b, type = "rstudent", quadratic=F, col = "lightblue",
             main = "After",
             ylim = c(-5, 5))
```

## New Residuals - Histogram comparison
```{r}
hist(studres(fit2), breaks = 25, freq=F, 
     col = "lightblue", 
     main = "Before",
     xlab = "Studentized Residuals",
     xlim = c(-4, 4))
hist(studres(fit4), breaks = 25, freq=F, 
     col = "lightblue", 
     main = "After",
     xlab = "Studentized Residuals",
     xlim = c(-4, 4))
```

## New Residuals - QQ plot comparison
```{r}
qqPlot(fit2, col = "grey", col.lines = "cornflowerblue",
       ylab = "Studentized Residuals",
       main = "Before",
       ylim = c(-4, 4))
qqPlot(fit4, col = "grey", col.lines = "cornflowerblue",
       ylab = "Studentized Residuals",
       main = "After",
       ylim = c(-4, 4))
```

## New Residuals - Standardized
```{r}
#Standardized residuals

barplot(height = stdres(fit4),
        main = "Standardized Residuals", xlab = "index", 
        ylab = "Standardized Residual", ylim = c(-4, 4))
abline(h = 3, col = "lightblue")
abline(h = -3, col = "lightblue")
```

## New Residuals - Studentized
```{r}
#Studentized residuals
barplot(height = studres(fit4),
        main = "Studentized Residuals", xlab = "index",
        ylab = "Studentized Residual", ylim = c(-4, 4))
abline(h = 3, col = "lightblue")
abline(h = -3, col = "lightblue")
```

## New Residuals - R Student
```{r}
#R student residuals
cor.qt <- qt(0.05/(2 * 48204), 48176, lower.tail = F)

barplot(height = rstudent(fit4),
        main = "R Student Residuals", xlab = "index",
        ylab = "R Student Residual", ylim = c(-6, 6))
abline(h=cor.qt, col = "lightblue", lwd=2)
abline(h=-cor.qt, col = "lightblue", lwd=2)
```

## New Residuals - Summary
```{r}
# Standardized
length(which(stdres(fit4) > 3))
length(which(stdres(fit4) < -3))

length(which(stdres(fit4) > 3.2))
length(which(stdres(fit4) < -3.2))

# Studentized
length(which(studres(fit4) > 3))
length(which(studres(fit4) < -3))

length(which(studres(fit4) > 3.2))
length(which(studres(fit4) < -3.2))

# R student
length(which(rstudent(fit4) > cor.qt))
length(which(rstudent(fit4) < -cor.qt))

# Total residuals
df_ResPost <- df_transformed[which(stdres(fit4) > 3 |
                                     stdres(fit4) < -3 |
                                     studres(fit4) > 3 |
                                     studres(fit4) < -3 |
                                     rstudent(fit4) > cor.qt |
                                     rstudent(fit4) < -cor.qt), ]

```

# Influence Analysis

## Influence - Measures
```{r}
inf <- influence.measures(fit4)
summary(inf)
influenceIndexPlot(fit4,vars=c("Cook"), 
                   main = "Cook's Distance")
influenceIndexPlot(fit4,vars=c("Hat"), 
                   main = "Hat Values")

```

## Influence - DFBETAS
```{r}
dfbetasPlots(fit4, intercept = F,
             terms = "holiday",
             main = "Holiday",
             col = "deepskyblue3")

dfbetasPlots(fit4, intercept = F,
             terms = "temp",
             col = "deepskyblue3")

dfbetasPlots(fit4, intercept = F,
             terms = "timeInterval",
             col = "deepskyblue3")

dfbetasPlots(fit4, intercept = F,
             terms = "weather_main",
             col = "deepskyblue3",
             layout = c(2, 2),
             ask = F,
             xlab = NA)
```

# Final Model

```{r}
summary(fit4)
```








